Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-02-13 paper/workshop/sparse_poly.tex
cleaned up intro

\begin{abstract}
We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
matrix of size $N \times D$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.

@@

Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
While their use has long been widespread, the authors are unaware of any improvements made to their calculation efficiency.
In this work we provide an algorithm for calculating polynomial features for a matrix of size matrix of size $NxD$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
The density of a matrix is the percent of its nonzero items, so $0 \le d \le 1$. The standard algorithm is $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^kN)$ as opposed to $O(D^kN)$.

\section{Algorithm}
The standard method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, such products will be common, and since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

@@

While the idea is straightforward, there is yet one unaddressed caveat; 
Given a set of nonzero columns whose product was just calculated to generate a polynomial feature, where in the polynomial matrix does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the form

>>>>

\begin{abstract}
We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{X}$.

@@

Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
In this work we provide an algorithm for calculating polynomial features for a row vector $\vec{x}$ of a $N \times D$ matrix with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
The density of a vector is the percent of elements that are nonzero, so $0 \le d \le 1$.
The standard algorithm has time and space complexity $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^kN)$ as opposed to $O(D^kN)$.

\section{Algorithm}
The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, products resulting in zero will be common.
Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

@@

While the idea is straightforward, there is yet one unaddressed caveat.
Given a set of nonzero columns whose corresponding components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the following form:
